The Chemistry Development Kit (CDK) \cdkversion{}: atom typing, depiction, molecular formulas, and substructure searching

### 2017-04-26 article.tex
Remove the SMARTS benchmark - we can summarise the same point with the more comprehensive benchmark.

CDK \cdkversion{} includes large improvements to algorithm
efficiency. A simple benchmark experiment demonstrates the relative
improvements between recent CDK versions and other open source
toolkits (Table~5\ref{tab:smigrep}). The efficiency improvements are a
combination of optimising data structures and key molecule processing
algorithms (e.g. kekulisation and aromaticity) needed before a SMARTS
match can be run.

@@

The results of benchmark are summarised in Table 6 and 7. The total elapsed
times are reported in Table 6, Table 7 subtracts the first tasks results 
(Count Heavy Atoms) to provide a comparable measure without the overhead of
input read time. The throughput as molecules per minute is reported but is less
accurate for v. fast running tasks.

@@

\textbf{-sssr}, The second subtask computes the Smallest Set of Smallest Rings 
(SSSR) and reports the size of the SSSR (circuit rank) for each record. Although 
circuit rank can be computed more efficiently 
with a linear traversal (counting DFS back-edges) or with Euler's 
polyhedron formula we are testing the time to enumerate the SSSR set. In general
SSSR is considered unfavourable due to the 
non-uniqueness of the set and need for Gaussian matrix elimination (cubic 
runtime). With some bookkeeping the time spent in the matrix elimination has been
reduced~\cite{May2014}. For ChEMBL we see the time to generate the SSSR is now
$\sim$16 seconds when it previously took around $\sim$3.5 minutes (Table 7).

@@

The speed-up in \cdkversion{} is slightly better than the SSSR task. ChEMBL 
previously spent 4-5 minutes and now takes only $\sim$14 seconds (Table 7).
In \cdkversion{} finding all rings ($\leq$ 12 bonds) runs faster than the 
non-unique SSSR computation.

@@

\textbf{-ofmt sdf}: For writing SDF output there is minimal improvement from v1.14.19, when 
discounting improvements in read performance the SDF generation for ChEBI 
actually runs slightly slower than v1.4.19 (Table 7). This can be partially 
explained by the more comprehensive SDF generation that now writes Sgroups 
as well as computing values for atom parity and valence columns.

@@

\subsection*{Table 5 - SMARTS matching performance.}\label{tab:smigrep}
  Total time taken to match the SMARTS \texttt{O=[C,N]aa[N,O;!H0]} against $\sim$1.5 million SMILES from
  ChEMBL 21. The reported time is for single core execution, measure with the Unix time utility (Cent OS 7, Intel Core i7-4790 CPU @ 3.60GHz). The times for recent versions of Open Babel~\cite{OBoyle2011} and RDKit~\cite{rdkit} are included for reference but this is not intended to be a systematic comparison. The difference in number found is due to variations in valence and aromaticity models.
  \vskip \baselineskip

\begin{tabular}{lll}
Toolkit                   & Time (s)   & Matched \\
\hline
CDK v1.4.9                 & 3,843  & 69,782  \\
CDK \cdkversion{}         & 67   & 62,518  \\
Open Babel 2.3.90         & 835 & 61,906  \\
RDKit 2016.3.4            & 378  & 62,402  \\
\end{tabular}

\subsection*{Table 6 - Summary of systematic benchmark comparing v1.4.19 to \cdkversion{}.}\label{tab:benchmark}
  The total elapsed \textit{real} time was measured with the unix time utility. The 
  throughput is reported in molecules per minute (K=thousand, M=million) as a 
  relatable metric.
  This throughput was calculated by taking the total elapsed
  time and dividing it by the number of molecule in the dataset (42704 for 
  ChEBI 149, and 1678393 for ChEMBL 22.1). The ChEBI SMILES input contains 
  2107 blank (but valid) inputs, this accounts for the majority skipped 
  in v1.4.19. The throughput calculation was adjust to account for this.

@@

\subsection*{Table 7 - Summary of systematic benchmark comparing v1.4.19 to \cdkversion{} without read times.}
\label{tab:benchmark-delta}
The number of records skipped and time to run the countheavy benchmark has been subtracted from Table 6. The remaining results provides a relative comparison without accounting for the overhead of reading the input.

>>>>

CDK \cdkversion{} includes large improvements to algorithm
efficiency. This is emphasised in the systematic benchmark
of MACCS-like 166 key generation (Table~5). The efficiency improvements
are a combination of optimising data structures and key molecule processing
algorithms (e.g. kekulisation and aromaticity) needed before a SMARTS
match can be run.

@@

The results of benchmark are summarised in Table 5 and 6. The total elapsed
times are reported in Table 5, Table 6 subtracts the first tasks results 
(Count Heavy Atoms) to provide a comparable measure without the overhead of
input read time. The throughput as molecules per minute is reported but is less
accurate for v. fast running tasks.

@@

\textbf{-sssr}, The second subtask computes the Smallest Set of Smallest Rings 
(SSSR) and reports the size of the SSSR (circuit rank) for each record. Although 
circuit rank can be computed more efficiently 
with a linear traversal (counting DFS back-edges) or with Euler's 
polyhedron formula we are testing the time to enumerate the SSSR set. In general
SSSR is considered unfavourable due to the 
non-uniqueness of the set and need for Gaussian matrix elimination (cubic 
runtime). With some bookkeeping the time spent in the matrix elimination has been
reduced~\cite{May2014}. For ChEMBL we see the time to generate the SSSR is now
$\sim$16 seconds when it previously took around $\sim$3.5 minutes (Table 6).

@@

The speed-up in \cdkversion{} is slightly better than the SSSR task. ChEMBL 
previously spent 4-5 minutes and now takes only $\sim$14 seconds (Table 6).
In \cdkversion{} finding all rings ($\leq$ 12 bonds) runs faster than the 
non-unique SSSR computation.

@@

\textbf{-ofmt sdf}: For writing SDF output there is minimal improvement from v1.14.19, when 
discounting improvements in read performance the SDF generation for ChEBI 
actually runs slightly slower than v1.4.19 (Table 6). This can be partially 
explained by the more comprehensive SDF generation that now writes Sgroups 
as well as computing values for atom parity and valence columns.

@@

\subsection*{Table 5 - Summary of systematic benchmark comparing v1.4.19 to \cdkversion{}.}\label{tab:benchmark}
  The total elapsed \textit{real} time was measured with the unix time utility. The 
  throughput is reported in molecules per minute (K=thousand, M=million) as a 
  relatable metric.
  This throughput was calculated by taking the total elapsed
  time and dividing it by the number of molecule in the dataset (42704 for 
  ChEBI 149, and 1678393 for ChEMBL 22.1). The ChEBI SMILES input contains 
  2107 blank (but valid) inputs, this accounts for the majority skipped 
  in v1.4.19. The throughput calculation was adjust to account for this.

@@

\subsection*{Table 6 - Summary of systematic benchmark comparing v1.4.19 to \cdkversion{} without read times.}
\label{tab:benchmark-delta}
The number of records skipped and time to run the countheavy benchmark has been subtracted from Table 6. The remaining results provides a relative comparison without accounting for the overhead of reading the input.
